---
title: Why plotthedata?
author: Ryan Chitwood
date: '2019-11-12'
slug: why-plotthedata
categories: []
tags: []
---

# Introduction

In data science, there's always some hip new technology, model, or package to try out. It's great to work in such a growing field, but sometimes I tire from all the newness. When that happens, I hearken back to the words of an old professor of mine:

> Always **plot the data**.

This pearl has followed me from my first semester of grad school to this day, so coming up with a name for my personal website was easy: plotthedata.

# Case Study: Anscombe's Quartet

When beginning any data analysis, it's easy to jump right to the statistics, but so much can be missed when you don't plot the data. A famous group of data sets exemplifies this folly: [Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet). And it's one of R's built-in data sets!

```{r, anscombe}
head(anscombe)
```

## Tidy

Unfortunately, `anscombe` breaks the first rule of tidy data:

> Each variable must have its own column.

Let's fix that. First, load the tidyvsere package:

```{r, setup, message = FALSE}
library(tidyverse)
```

Now, tidy:

```{r, anscombe_tidy}
anscombe_tidy <- anscombe %>% 
  mutate(observation = row_number()) %>% 
  gather(key, value, -observation) %>% 
  separate(key, into = c("var", "set"), sep = 1) %>% 
  spread(var, value)

head(anscombe_tidy)
```

## Summary Statistics

Phew, that's better. Now, let's compute some summary statistics:

```{r, set_means}
anscombe_tidy %>% 
  group_by(set) %>% 
  summarize(
    mean(x),
    var(x),
    mean(y),
    var(y),
    cor(x, y)
  )
```

The summary statistics are nearly identical! That's weird. Let's run some models to see what's going on here.

## Model

Each data set has a single continuous explanatory variable and single continuous response variable. Sounds like a job for simple linear regression! Now, we *could* run four separate regressions, but that's no fun. This is R! We have the power to do better! Let's use the power of `tidyr::nest()`:

```{r, nest_by_set}
by_set <- anscombe_tidy %>% 
  group_by(set) %>% 
  nest()
```

So now we have a grouped data frame where each row is an Anscombe set and the data column contains that data set. Pretty neat! Now let's create our model function:

```{r, set_model}
set_model <- function(df) {
  lm(y ~ x, data = df)
} 
```

Now, let's map the model function to our nested data frame:

```{r, map_model}
by_set <- by_set %>% 
  mutate(model = map(data, set_model))
```

We now have a columns that contains a list of our data and a list of our models. Let's use the broom package to extract some of the parameter estimates:

```{r, tidy_fit}
library(broom)

by_set %>% 
  mutate(fit = map(model, tidy)) %>% 
  select(set, fit) %>% 
  unnest(fit) %>% 
  arrange(term, set)
```

Wow! We can see clearly that the summary statistics *and* model summaries are nearly identical for all data sets. This is bananas. 

## Plot

Okay, we've avoided it long enough. It's time to plot the data.

```{r, plot}
anscombe_tidy %>% 
  ggplot(aes(x, y)) + 
  geom_point() + 
  facet_wrap(~set) + 
  geom_smooth(method = "lm", se = FALSE)
```

And the curtain is lifted. We can see clearly that these are wildly different data sets! We would be lost on all accounts if we neglected to **plot the data**.

# Conclusion

This has been a contrived example, but it serves to prove a larger point: the fundamentals are important. Next time you are staring down the chasm between your current skill set and where you want to be, do me a favor: stop, take a deep breath, and **plot the data**.
