---
title: Why plot the data?
author: Ryan Chitwood
date: '2020-01-01'
publishdate: '2020-01-01'
slug: why-plot-the-data
categories: []
tags: []
---

In data science there's always some new technology to try out. It's great to work in such a growing field, but sometimes it feels hard to keep up. When that happens, I remember the first step of any data analysis:

> Plot the data.

This was the refrain of one of my mentor professors in grad school, and it has stuck with me ever since. Dreaming up a name for my personal website was easy: plot the data.

# Case Study: Anscombe's Quartet

Let's look at a classic example. [Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe's_quartet) was created to expose the pitfall of not plotting the data. And it's one of R's built-in data sets!

```{r, anscombe}
head(anscombe)
```

## Tidy

Unfortunately, `anscombe` breaks the first rule of tidy data:

> Each variable must have its own column.

Let's fix that. First, load the tidyverse package:

```{r, tidyverse, message = FALSE}
library(tidyverse)
```

Now, tidy:

```{r, anscombe_tidy}
anscombe_tidy <- anscombe %>% 
  pivot_longer(everything(),
    names_to = c(".value", "set"),
    names_pattern = "(.)(.)"
  ) %>%
  arrange(set)
anscombe_tidy
```

## Summary Statistics

Phew, that's better. Now, let's compute some summary statistics:

```{r, set_means}
anscombe_tidy %>% 
  group_by(set) %>% 
  summarize(
    mean(x),
    var(x),
    mean(y),
    var(y),
    cor(x, y)
  )
```

The summary statistics are nearly identical! That's weird. Let's run some models to see what's going on here.

## Model

Each data set has one continuous explanatory variable and one continuous response variable. Sounds like a job for simple linear regression! Now, we *could* run four separate regressions, but that's no fun. This is R! We can do better! Let's use the power of the tidymodels package:

```{r, tidymodels, message = FALSE}
library(tidymodels)
```

```{r, nest_by_set}
by_set <- anscombe_tidy %>% 
  group_by(set) %>% 
  nest()
```

So now we have a grouped data frame where each row is an Anscombe set and the data column contains that data set. Pretty neat! Now let's create our model function:

```{r, set_model}
set_model <- function(df) {
  lm(y ~ x, data = df)
} 
```

Now, let's map the model function to our nested data frame:

```{r, map_model}
by_set <- by_set %>% 
  mutate(fit = map(data, set_model))
```

We now have a columns that contains a list of our data and a list of our models. Let's use the broom package to extract some of the parameter estimates:

```{r, tidy_fit}
by_set %>% 
  mutate(tidied = map(fit, tidy)) %>% 
  select(set, tidied) %>% 
  unnest(tidied) %>% 
  arrange(term, set)
```

Wow! We can see clearly that the summary statistics *and* model summaries are nearly identical for all data sets. This is crazy!

## Plot

Okay, we've put it off long enough. It's time to plot the data.

```{r, plot}
anscombe_tidy %>% 
  ggplot(aes(x, y)) + 
  geom_point() + 
  facet_wrap(~set) + 
  geom_smooth(method = "lm", formula = y ~ x)
```

And the curtain is lifted. We can clearly see that each set is quite different:

* Set 1 is a linear relationship with some variance.
* Set 2 is curvilinear.
* An outlier messes up a linear relationship in set 3.
* Another outlier in set 4 creates a relationship when there isn't one.

# Conclusion

The jig is up. This is a contrived example, but it proves a larger point: don't neglect the fundamentals. Next time you are starting an analysis, do yourself a favor and don't forget to **plot the data**.
