---
title: "Probability Distributions and Their Stories: Baller Edition"
date: "2024-08-01"
---

Can you ball with the best (probability distributions)?
ðŸ€

# Fun with distributions

The first time I ever learned about probability I was in math class at [St. Elmo](https://www.muscogee.k12.ga.us/about-us-clone/families-students/teaching-and-learning/advanced-learning-programs/gifted).
I love math, and I loved going to St. Elmo once a week as a kid.
There we were, a bunch of 4th- and 5th-graders ready to learn about games of chance and predicting the future!
Except no.
The teacher gave us bags of marbles.

> Each bag has 5 white marbles and 5 red ones.
> Draw a marble and don't put it back.
> What are the chances that the next marble is the same color?

Lame.
So here I am years later righting this wrong (don't blame the teacher; blame the curriculum).
We're going to learn about balling with probability distributions.
Basketball that is.

NB: This entire post is based on the excellent tutorial, [Probability distributions and their stories](http://bebi103.caltech.edu.s3-website-us-east-1.amazonaws.com/2018/tutorials/t3c_probability_stories.html), by [Justin Bois](http://bois.caltech.edu/).
My aim is to translate Python code shown there to R using the {[distributions3](https://alexpghayes.github.io/distributions3/)} package and to come up with my own basketball examples for each distribution.

# What is a probability distribution?

As usual, Wikipedia has [the definition](https://en.wikipedia.org/wiki/Probability_distribution):

> A probability distribution is the mathematical function that gives the probabilities of occurrence of possible outcomes for an experiment.

In my words, it's a mathematical description of the chances that an event happens.
Probability distributions can describe discrete (typically countable) or continuous (typically measured) variables.
Let's dive in.

## The distributions3 package

The {[distributions3](https://alexpghayes.github.io/distributions3/)} package a provides a number of probability distributions as S3 objects.
It also provides generic functions that work with each distribution.
It's a really nice and tidy API.
I think base R's probability functions should probably have been designed this way from the start.

The main generics are:

-   `random()`: Draw samples from a distribution.
-   `pdf()`: Evaluate the probability density (or mass) at a point.
-   `cdf()`: Evaluate the cumulative probability up to a point.
-   `quantile()`: Determine the quantile for a given probability. Inverse of `cdf()`.

Let's load the package and go from there.

```{r}
library(distributions3)
```

# Discrete distributions

## Bernoulli distribution

A *Bernoulli trial* is an experiment that has two outcomes that can be encoded as success ($y=1$) or failure ($y=0$).
The result $y$ of a Bernoulli trial is Bernoulli distributed.

[Elena Delle Donne](https://en.wikipedia.org/wiki/Elena_Delle_Donne), the greatest free throw shooter of all time, has a career free throw success rate of 93.4%.
The probability that she makes her next free throw is Bernoulli distributed.

```{r}
X <- Bernoulli(0.7)
X

mean(X)
variance(X)
skewness(X)
kurtosis(X)

random(X, 10)
pdf(X, 1)
log_pdf(X, 1)
cdf(X, 0)
quantile(X, 0.7)

cdf(X, quantile(X, 0.7))
quantile(X, cdf(X, 0.7))
```

## Binomial

What if we are interested in more than one shot?
It turns out that the Bernoulli distribution is just a special case of the Binomial distribution where $N=1$.

If we perform $N$ Bernoulli trials, each with probability $\theta$ of success.
The number of successes, $n$, is Binomially distributed.

[Shaquille O'Neal](https://en.wikipedia.org/wiki/Shaquille_O'Neal) will always be one of the most dominant big men of all time, but his free throw percentage, by comparison was infamously atrocious.
He averaged just 52.7% from the line with an average of 9.3 free-throw attempts per game.

```{r}
X <- Binomial(size = 10, p = 0.527)
X

mean(X)
variance(X)
skewness(X)
kurtosis(X)

random(X, 10)

pdf(X, 2L)
log_pdf(X, 2L)

cdf(X, 4L)
quantile(X, 0.7)

cdf(X, quantile(X, 0.7))
quantile(X, cdf(X, 7))
```
